Patterns of stress tolerance versus resistance in genome-wide expression data

### 2014-07-16 doc/ApTranscriptome_manuscript.Rmd

---
title: "Pattners of stress tolerance versus resistance in genome-wide expression data
  of sympatric ant species"
author: "John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn,
  Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan"
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
  html_document: null
  word_document: default
bibliography: ApTranscriptome.bib
---

@@

In this study, we apply a novel method to characterize the reaction-norm of all genes in the transcriptome (e.g. the reactionome) in response to thermal stress to overcome the above limitations. Specifically, we measure gene expression across a thermal gradient and identify thermally-responsive genes using a regression approach. Instead of focusing on individual genes or groups of genes with differential expression, we compare overall patterns in the reactionomes between closely-related species to gain insight into how spatially-divergent selection has changed gene expression patterns.

We apply this approach in a study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively. Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguezcabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006]. As ectothermic species, ants experience a wide range of temperature conditions. Throughout the year, ant species persist through freezings and hot summers, and at shorter time scales, individual workers can experience substantial thermal shock as they leave the buffered ant nest to forage for food. Thus, at the organismal level we expect that ants utilize multiple molecular pathways to tolerate thermal stress.

To explore the molecular toolkit used by *Aphaenogaster* to cope with thermal stress, we performed a whole organism RNA sequencing study. Our specific goals were to 1) generate the set of all genes (i.e. transcriptome) expressed by *Aphaenogaster* across physiologically-relevant temperatures, 2) identify genes that have responsive expression across this thermal gradient (i.e. thermal reactionome), 3) explore the functional annotations of genes that are activated or down-regulated in response to cold and high temperatures. We expected to observe candidate genes in the *response to stress* Gene Ontology (GO) category including *heat shock proteins* and ... in the thermally-responsive group.

@@

Finally, to test whether the temperature at which thremally-responsive genes are activated differs between species, we identified the temperature at which there was the greatest change in expression. We grouped the *High* genes along with the high temperature end of the *Bimodal* genes, and did the same for *Low* genes. We then tested if the temperature at which the *High* or *Low* group differed among the two species.

@@

```{r mosaic_plot, echo=FALSE, results='hide'}
tiff("mosaic_plot.tiff")
mosaic.colors <- brewer.pal(5, "Blues")
par(mar = c(5.1, 3.1, 2.1, 7.5))
mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
labels <- c("Low", "Intermediate", "High", "Bimodal", "NotResp")
text(c(.1,.5,.75,.85,.99), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
text(.5, par("usr")[4]+.1, "A. carolinensis", xpd = TRUE, font = 3)
text(.2, .6, "2,594")
text(.85, .03, "223", cex = .75)
legend(1.05, .9, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
dev.off()
```

>>>>

---
title: "Patterns of stress tolerance versus resistance in genome-wide expression data
  of sympatric ant species"
author: "John Stanton-Geddes, Andrew Nguyen, Mahesh Vangala, Jim Vincent, Rob Dunn,
  Aaron Ellison, Nate Sanders, Nick Gotelli, Sara Helms-Cahan"
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex
  html_document: null
  word_document: default
bibliography: ApTranscriptome.bib
---

@@

In this study, we apply a novel method to characterize the reaction norm of all genes in the transcriptome (e.g. the reactionome) in response to thermal extremes to overcome the above limitations. Specifically, we measure gene expression across a thermal gradient and identify thermally-responsive genes using a regression approach. Instead of focusing on individual genes or groups of genes with differential expression, we compare overall patterns in the reactionomes between closely-related species to gain insight into how spatially divergent selection has changed patterns of gene expression.

We apply this approach in a study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively. Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguezcabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006]. Temperature is a major selective pressure as *Aphaenogaster* colonies experience freezing winters and hot summers throughout the year, and at shorter time scales, individual workers can experience large thermal changes as they leave the buffered ant nest to forage for food. *A. carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A. picea* [cite...]. Thus, we expect patterns of gene expression underlying tolerance of thermal extremes to differ between these species.

To explore differences in the thermal reactionomes between *A. picea* and *A. carolinensis*, we performed a whole organism RNA sequencing study spanning temperatures from 0° to 38.5°C. We identified genes that have responsive expression across this thermal gradient (i.e. thermal reactionome), compared the patterns

3) explore the functional annotations of genes that are activated or down-regulated in response to cold and high temperatures. We expected to observe candidate genes in the *response to stress* Gene Ontology (GO) category including *heat shock proteins* and ... in the thermally-responsive group.

@@

Finally, to test whether the temperature at which thermally-responsive genes are activated differs between species, we identified the temperature at which there was the greatest change in expression. We grouped the *High* genes along with the high temperature end of the *Bimodal* genes, and did the same for *Low* genes. We then tested if the temperature at which the *High* or *Low* group differed among the two species.

@@

```{r mosaic_plot, echo=FALSE, results='hide'}
tiff("mosaic_plot.tiff")
mosaic.colors <- brewer.pal(5, "Blues")
par(mar = c(5.1, 3.1, 2.1, 7.5))
mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
labels <- c("Low", "Intermediate", "High", "Bimodal", "NotResp")
text(c(.1,.5,.75,.85,.99), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
text(.5, par("usr")[4]+.1, "A. carolinensis", xpd = TRUE, font = 3)
text(.15, .7, "2,594")
text(.85, .03, "223", cex = .75)
legend(1.05, .9, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
dev.off()
```
